Complex Analysis of Askaryan Radiation: UHE-$\nu$ Identification and Reconstruction via the Hilbert Envelope of Observed Signals

### 2025-09-18 template.tex
grinding towards the final results in Table III

\begin{abstract}
The detection of ultra-high energy neutrinos (UHE-$\nu$), with enegies above 10 PeV, has been a long-time goal in astroparticle physics.  Autonomous, radio-frequency (RF) UHE-$\nu$ detetectors have been deployed in polar regions.  These detectors rely on the Askaryan effect in ice for the neutrino signal.  The Askaryan effect occurs when the excess negative charge within a high-energy cascade radiates in a dense medium.  UHE-$\nu$ can induce such cascades that radiate in the RF bandwidth above thermal backgrounds.  To identify UHE-$\nu$ signals in future data from Askaryan-class detectors, analytic models of the Askaryan electromagnetic field have been created and compared to simulations and laboratory measurements.  These models have correctly described the Askaryan electromagnetic field, while leaving the effect of the RF detection channels on the field to simulation packages.  In this work, we present a fully analytic Askaryan model that accounts for the effect of an RF detection channel.  First, we derive formulas for the observed voltage trace, and for the Hilbert envelope of the trace.  Second, we match the analytic model to data computed for a detector with a single string of eight RF detection channels in NuRadioMC, a key Monte Carlo toolset in UHE-$\nu$ detection.  For UHE-$\nu$ signals above RF thermal backgrounds, we find correlation coefficients in excess of $0.95$.  Third, 96.9\% of UHE-$\nu$ signals pass a correlation threshold of 0.386, while all but 0.877 out of $1.57 \times 10^8$ RF thermal triggers are rejected.  At a thermal trigger rate of 1 Hz, this corresponds to 5 years of continuous operation.  Finally, we reconstruct the logarithm of the UHE-$\nu$ cascade energy.
\end{abstract}

@@

\begin{itemize}
\item The result for $\mathcal{E}_{r*s}(t)$, Eqs. \ref{eq:final} and \ref{eq:final2}, depends on the model for $s(t)$, Eq. \ref{eq:s}.  Equation \ref{eq:s} is a simplified version of Eq. 28 in the analysis presented by Hanson and Hartig (HH) \cite{PhysRevD.105.123019}.  The $E_0$ simply represents all time-independent amplitude factors.  The full expression for $s(t)$ is
\begin{equation}
r\vec{E}(t_{\rm r},\theta) = -\frac{E_0\omega_0 \sin(\theta)}{8\pi p}t_{\rm r}e^{-\frac{t_{\rm r}^2}{4p}+p\omega_0^2}\erfc(\sqrt{p}\omega_0) \label{eq:s_full}
\end{equation}
\begin{table*}
\begin{tabular}{| c | c | c |} \hline
\textbf{Variable} & \textbf{Definition} & \textbf{Units}\\ \hline
$c$ & speed of light in medium & m ns$^{-1}$ \\ 
$r$ & distance to cascade peak & m \\
$t_{\rm r}$ & $t-r/c$ & ns \\
$\theta_{\rm C}$ & Cherenkov angle & radians \\ 
$\theta$ & viewing angle from cascade axis & radians \\ 
$a$ & longitudinal cascade length (see \cite{10.1103/physrevd.65.016003}) & m \\ 
$n_{max}$ & max excess cascade particles (see \cite{10.1103/physrevd.65.016003})  & none \\
$E_{\rm 0}$ & $\propto n_{\rm max}a$ (see \cite{10.1103/physrevd.65.016003}) & V GHz$^{-2}$ \\
$p$ & $\frac{1}{2}(a/c)^2 \left(\cos\theta - \cos\theta_C\right)^2$ (see \cite{PhysRevD.105.123019}) & ns$^2$ \\ 
$\omega_{\rm 0}$ & $\sqrt{\frac{2}{3}} (c\sqrt{2\pi}\rho_{\rm 0})/(\sin\theta)$ (see \cite{10.1016/j.astropartphys.2017.03.008}) & GHz \\
$\sqrt{2\pi}\rho_{\rm 0}$ & lateral ICD width (see \cite{10.1016/j.astropartphys.2017.03.008}) & m$^{-1}$ \\ \hline
\end{tabular}
\caption{\label{tab:param} things.}
\end{table*}

@@

In Fig. \ref{fig:fig3}, the circles represent the normalized histogram of the correlation coefficent between the optimized analytic envelope and thermal noise.  A fitting function of the form $x^2 \exp(-0.5 x^2)$ was fit to the noise distribution, and is represented by the gray dashed line.  The solid gray line represents the signal distribution, which peaks at a correlation value of 0.94.  Lower signal correlation values correspond to lower SNR values (Fig. \ref{fig:fig4}).  The vertical black dashed line represents a threshold of 0.4.  For the simulated UHE-$\nu$, 99.99\% of correlations between CSWs and analytic envelopes are greater than or equal to this threshold.  Assuming a thermal trigger rate of 1 Hz, integrating the PDF of the noise disteribution above the correlation threshold is equivalent to 0.2 noise events every 5 years.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,trim=3.25cm 8.25cm 4.5cm 9.0cm,clip=true]{Aug18_plot1.pdf}
\caption{\label{fig:fig5} (Black circles) Viewing angle from NuRadioMC. (Solid black line) Gaussian fit, with $\mu = 55.6$ deg, and $\sigma = 5.77$ deg.  (Black dashed line) Cherenkov angle, $\theta_{\rm C}$.}
\end{figure}

\item Equations \ref{eq:uncert}-\ref{eq:a_err} may be used to reconstruct the natural logarithm of the UHE-$\nu$ cascade energy, $\ln\Lambda$.  For Eq. \ref{eq:uncert}, $\sigma_t$ is measured from the optimized analytic envelope, $c$ and $\theta_{\rm C}$ are known constants, and an assumption must be made for $\Delta\theta$.  We will make the assumption that $\Delta\theta \approx \Delta\theta_{\rm rms}$.  Relying on a single string of RF channels makes calculating $\Delta\theta$ difficult.  However, the assumption is justified since the $\Delta\theta$ distribution is Gaussian.  The MC truth for $\Delta\theta$ in this analysis is shown in Fig. \ref{fig:fig5}, with a Gaussian fit ($\mu=55.6$ deg, $\sigma=5.77$ deg).  The rms and the $\sigma$ parameter are identical for Gaussian distributions, so we assume $\Delta\theta \approx 5.77$ deg.  The fractional error $\sigma_{\Delta\theta}/\Delta\theta$ is then set to 1.0, reflecting uncertainty in this parameter.  Solving Eq. \ref{eq:uncert} for $a$ gives

@@

\begin{equation}
\log_{10} E_{\rm C} = \frac{\left( \frac{c\sigma_t}{c_{\rm em/had} \Delta\theta_{\rm rms}\sin\theta_{\rm C}} \right)^2}{\ln 10} + \log_{10} E_{\rm crit} \label{eq:lnLambda_2}
\end{equation}

Figure \ref{fig:fig6} contains the $\sigma_t$ distribution, converted into $\log_{10} E_{\rm C}$.  Note that when applying the change of base formula, factors of $\ln(10)$ cancel in the ratio of Eq. \ref{eq:a_err_2}.  The histogram is spread out over a range much larger than the expected range from NuRadioMC.  However, the average value of the histogram is $15.7 \pm 4.5$, in agreement with the prediction from NuRadioMC and Eq. \ref{eq:a_err_2}: $16.7 \pm 2.0$.  Thus, the energy reconstruction produces the correct order of magnitude on average, if not the precise value.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,trim=3.25cm 8.25cm 4.5cm 8.25cm,clip=true]{Aug19_plot1.pdf}
\caption{\label{fig:fig6} Normalized distribution of $\sigma_t$ values, converted to $\log_{10}E_{\rm C}$ via Eq. \ref{eq:lnLambda_2}.  The vertical black line represents the average $\log_{10}E_{\rm C}$ from the NuRadioMC energies, and the shaded region represnts the $1\sigma$ error range using Eq. \ref{eq:a_err_2}.}
\end{figure}

>>>>

\begin{abstract}
The detection of ultra-high energy neutrinos (UHE-$\nu$), with enegies above 10 PeV, has been a long-time goal in astroparticle physics.  Autonomous, radio-frequency (RF) UHE-$\nu$ detetectors have been deployed in polar regions.  These detectors rely on the Askaryan effect in ice for the neutrino signal.  The Askaryan effect occurs when the excess negative charge within a high-energy cascade radiates in a dense medium.  UHE-$\nu$ can induce such cascades that radiate in the RF bandwidth above thermal backgrounds.  To identify UHE-$\nu$ signals in future data from Askaryan-class detectors, analytic models of the Askaryan electromagnetic field have been created and compared to simulations and laboratory measurements.  These models have correctly described the Askaryan electromagnetic field, while leaving the effect of the RF detection channels on the field to simulation packages.  In this work, we present a fully analytic Askaryan model that accounts for the effect of an RF detection channel.  First, we derive formulas for the observed voltage trace, and for the Hilbert envelope of the trace.  Second, we match the analytic model to data computed for a detector with a single string of eight RF detection channels in NuRadioMC, a key Monte Carlo toolset in UHE-$\nu$ detection.  For UHE-$\nu$ signals, we find correlation coefficients between the analytic signal envelope and MC data in excess of $0.94$, and 99.99\% of UHE-$\nu$ signals pass a correlation threshold of 0.4.  Integrating the normalized PDF of the correlation distribution for RF thermal noise reveals that just 0.2 thermal background events pass the correlation threshold in 5 years of continuous operation at a 1 Hz thermal trigger rate.  Finally, we reconstruct the logarithm of the UHE-$\nu$ cascade energy.
\end{abstract}

@@

\begin{itemize}
\item The result for $\mathcal{E}_{r*s}(t)$, Eqs. \ref{eq:final} and \ref{eq:final2}, depends on the model for $s(t)$, Eq. \ref{eq:s}.  Equation \ref{eq:s} is a simplified version of Eq. 28 in the analysis presented by Hanson and Hartig (HH) \cite{PhysRevD.105.123019}.  The $E_0$ simply represents all time-independent amplitude factors.  The full expression for $s(t)$ is
\begin{equation}
r\vec{E}(t_{\rm r},\theta) = -\frac{E_0\omega_0 \sin(\theta)}{8\pi p}t_{\rm r}e^{-\frac{t_{\rm r}^2}{4p}+p\omega_0^2}\erfc(\sqrt{p}\omega_0) \label{eq:s_full}
\end{equation}
\begin{table*}
\begin{tabular}{| c | c | c |} \hline
\textbf{Variable} & \textbf{Definition} & \textbf{Units}\\ \hline
$c$ & speed of light in medium & m ns$^{-1}$ \\ 
$r$ & distance to cascade peak & m \\
$t_{\rm r}$ & $t-r/c$ & ns \\
$\theta_{\rm C}$ & Cherenkov angle & radians \\ 
$\theta$ & viewing angle from cascade axis & radians \\ 
$a$ & longitudinal cascade length (see \cite{10.1103/physrevd.65.016003}) & m \\ 
$n_{max}$ & max excess cascade particles (see \cite{10.1103/physrevd.65.016003})  & none \\
$E_{\rm 0}$ & $\propto n_{\rm max}a$ (see \cite{10.1103/physrevd.65.016003}) & V GHz$^{-2}$ \\
$p$ & $\frac{1}{2}(a/c)^2 \left(\cos\theta - \cos\theta_C\right)^2$ (see \cite{PhysRevD.105.123019}) & ns$^2$ \\ 
$\omega_{\rm 0}$ & $\sqrt{\frac{2}{3}} (c\sqrt{2\pi}\rho_{\rm 0})/(\sin\theta)$ (see \cite{10.1016/j.astropartphys.2017.03.008}) & GHz \\
$\sqrt{2\pi}\rho_{\rm 0}$ & lateral ICD width (see \cite{10.1016/j.astropartphys.2017.03.008}) & m$^{-1}$ \\ \hline
\end{tabular}
\caption{\label{tab:param} Parameters relevant for Eq. \ref{eq:s_full}.}
\end{table*}

@@

In Fig. \ref{fig:fig3}, the circles represent the normalized histogram of the correlation coefficent between the optimized analytic envelope and thermal noise.  A fitting function of the form $x^2 \exp(-0.5 x^2)$ was fit to the noise distribution, and is represented by the gray dashed line.  The solid gray line represents the signal distribution, which peaks at a correlation value of 0.94.  Lower signal correlation values correspond to lower SNR values (Fig. \ref{fig:fig4}).  The vertical black dashed line represents a threshold of 0.4.  For the simulated UHE-$\nu$, 99.99\% of correlations between CSWs and analytic envelopes are greater than or equal to this threshold.  Assuming a thermal trigger rate of 1 Hz, integrating the PDF of the noise distribution above the correlation threshold is equivalent to 0.2 noise events every 5 years.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth,trim=3.25cm 8.25cm 4.5cm 9.0cm,clip=true]{Aug18_plot1.pdf}
\caption{\label{fig:fig5} (Black circles) Viewing angle from NuRadioMC. (Solid black line) Gaussian fit, with $\mu = 54.8$ deg, and $\sigma = 2.96$ deg.  (Black dashed line) Cherenkov angle, $\theta_{\rm C}$.}
\end{figure}

\item Equations \ref{eq:uncert}-\ref{eq:a_err} may be used to reconstruct the natural logarithm of the UHE-$\nu$ cascade energy, $\ln\Lambda$.  For Eq. \ref{eq:uncert}, $\sigma_t$ is measured from the optimized analytic envelope, $c$ and $\theta_{\rm C}$ are known constants, and an assumption must be made for $\Delta\theta$.  We will make the assumption that $\Delta\theta \approx \Delta\theta_{\rm rms}$.  The assumption is justified since the $\Delta\theta$ distribution is Gaussian.  The MC truth for $\Delta\theta$ in this analysis is shown in Fig. \ref{fig:fig5}, with a Gaussian fit ($\mu=54.8$ deg, $\sigma=2.96$ deg).  The rms and the $\sigma$ parameter are equal for Gaussian distributions with zero mean, so we let $\Delta\theta \to \Delta \theta_{\rm rms}$.  The fractional error $\sigma_{\Delta\theta}/\Delta\theta$ is then set to 1.0, reflecting uncertainty in this parameter.  Solving Eq. \ref{eq:uncert} for $a$ gives

@@

\begin{equation}
\log_{10} E_{\rm C} = \frac{\left( c\sigma_t \right)^2}{\ln 10(c_{\rm em/had} \Delta\theta_{\rm rms}\sin\theta_{\rm C})^2} + \log_{10} E_{\rm crit} \label{eq:lnLambda_2}
\end{equation}

Table \ref{tab:2} contains the results for the energy reconstruction.  Note that when applying the change of base formula, factors of $\ln(10)$ cancel in the error ratio of Eq. \ref{eq:a_err_2}, so the error in $\log_{10} E_{\rm C}$ remains $2.0$.

\begin{table}
\centering
\begin{tabular}{| c | c |}
\hline
\textbf{MC Truth $\log_{10} E_{\rm C}$} & \textbf{Reconstructed $\log_{10} E_{\rm C}$} \\
17.0 & $17.8 \pm 2.0$ \\
\hline
\end{tabular}
\caption{\label{tab:2}}
\end{table}
